### 2018-01-23 ident/docs/ident_manuscript.tex
complete results and not so good discussion with new erfs

\begin{document}
\paragraph{Outline:}
\begin{itemize}
	\item Importance of parameter values for estimating in vivo response using kinetic models of metabolism and consequently for metabolic network design using kinetic models of metabolism
	\item The need for parameter identifiability to determine unique and true parameter values from observed data
	\item Types and purpose of identifiability for parameters
	\item Methods for structural identifiability and existing methods for practical identifiability
	\item Lack of methods for practical identifiability and consequently experimental design (not covered?)
	\item work done in this paper for practical identifiability			
	\item scalability of computer algebra-based methods for structural identifiability (using CRNT to reduce networks to make structural identifiability scalable) (move to discussion - may be in discussion)
\end{itemize}
\section{Introduction:}
Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks. In these models, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
\begin{subequations}\label{eq:kinstoich}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = f(x, \theta, u)
	\end{align}
\end{subequations}
The expression for the nonlinear function ($f$) used to describe each reaction flux $v_i$ in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction \parencite{Heijnen2005, Link2014, Machado2011, Srinivasan2015}. Accordingly, $f$ is a nonlinear function of the metabolite concentrations, enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

@@

In this paper, we propose a scalable methodology to establish practical identifiability for parameters in kinetic models of metabolism using abundantly available steady state concentration and flux data. We present a computer algebra-based method that can facilitate experimental design through practical identifiability of parameters separately for each individual reaction within a metabolic network based on available steady state experimental data. We illustrate the utility of this method by applying it to a small network of gluconeogenesis in \textit{E. coli} and demonstrating our ability to propose experiments that will facilitate parameter estimation for a kinetic model of this network. We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.

@@

The minimum number of experiments from which data is required for identifying all the parameters of a given flux is determined by dimension $\mathbb{R}^p$ of the parameter space of a chosen flux $v_i$. For instance, as demonstrated above, data from two distinct experiments is required for identifying the two parameters of $v_1$. In this case $p=2$. This also applies for identifying parameters in $v_2$.

Looking at the three fluxes whose parameters have to be identified, $v_3$ has the maximum number of parameters at three. Hence, we choose to select multiple distinct data sets, each with three different experiments to establish the practical identifiability of all the parameters in the model. Accordingly, we can select 4896 (18 x 17 x 16) distinct combinations of three experiments from the 18 different experiments mentioned earlier and elaborated in the Appendix.

\subsection{Experimental design through practical parameter identifiability for kinetic models of metabolism:}
Following the methodology described in Section \ref{sec:ident}, and demonstrated in Section \ref{sec:example} for a single flux using data from a combination of two different experiments, all distinct combinations found from Section \ref{sec:experiments} can be tested for their ability to practically identify parameters for any of the three fluxes in the small metabolic network. This step would help distinguish identifiable experiment combinations from combinations that do not practically identify any parameter in the model (Figure).
Subsequently, it is possible to obtain a collection of experiments that make up all identifiable data combinations that can be performed to obtain the most minimal and informative set of experiments to identify as many parameters as possible. Consequently, the set of experiments can be used to estimate the identifiable parameters in the model.

\section{Results:}
First in Section \ref{sec:closedform}, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. Then, in Section \ref{sec:trends} we discuss some of the trends we observed in the identifiability of different types of parameters using in silico experimental data from different parameter perturbations.%Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.
\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
Establishing parameter identifiability for fluxes in kinetic models of metabolism using the methodology we describe in the Methods section is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. We demonstrate one example of how the methodology works in Section \ref{sec:example} for parameters of flux $v_1$. The expression in Equation (\ref{eq:v1_par}) is obtained by using a computer algebra system. To recall, in this specific case, a computer algebra system is used to solve for the parameters of a flux described using the Michaelis-Menten kinetic rate law when experimental data from two different experiments is available.

However, we find that the nonlinearity of the MWC kinetic rate law used to model the allosteric regulation of $v_3$ makes it computationally intractable for determining the closed form expressions of the three parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ using a computer algebra system (Mathematica and SymPy in Python). In order to overcome this computational obstacle, we model the reaction rate for $v_3$ using the convenience kinetic rate law formulation (reference for convenience kinetics). The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}).

@@

This may not be a unique problem isolated to the system that we are investigating. Complicated expressions for describing fluxes have been extensively used to model observed experimental data for different fluxes in a variety of organisms (Chassagnole, other yeast and e. coli kinetic modeling papers). However, the identifiability of the parameters used in these models has never been truly examined. If these models are tractable, we believe our methodology can help in elucidating the identifiability of parameters in these models. If not, metabolic network fluxes can be expressed using alternative kinetic rate law models whose parameters can be tested for identifiability, and subsequently experiments can be designed for their estimation.

%Any test for identifiability is forgone in favor of direct parameter estimation using a nonlinear least squares optimization approach with powerful computers to overcome difficulties associated with getting global optimal solutions (). This is however paving way for probabilistic approaches based on Bayesian techniques (Lars Nielsen papers). Furthermore, none of these methods address the issue of designing experiments to suit parameter estimation. Parameter estimation for kinetic models has always focused on the ability to estimate parameters from existing data without the need for additional experiments, which might not be always possible if parameters are not identifiable from existing experimental data.

\subsection{Maximum reaction rates are more identifiable than enzyme binding constants:\label{sec:trends}
We use the percentage of data combinations that can identify a parameter as a simple measure of the degree of identifiability of that parameter. As an example, if 90\% of all the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%, and $\theta_i$ has a higher degree of identifiability than $\theta_j$. We can use this criteria to distinguish parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/noise_no_noise_kotte_ident_jan01}}
	\caption{The number of data sets, made from a combination of 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. In the presence of additive noise in the flux and concentration data used for identifiability, the number of data sets practically identifying parameters in fluxes d) $v_1$, e) $v_2$ and f) $v_3$ are shown. 25 samples of additive noise $\epsilon ~ N(0, 0.05)$ were used to generate noisy data for figures in the right hand side panel. The average and standard deviation values shown for identifiability of parameters with data in the presence of noise were calculated on the basis of the number of data combinations that can identify the respective parameter for each of the 25 noisy data samples.}\label{fig:figure1}
\end{figure}

In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. In the left panels (Figure \ref{fig:figure1}a-c), the data is shown when noise-free data is used for identification. The identifiability information for parameters using data with additive noise is shown on the right panel (Figure \ref{fig:figure1}d-f). The data shown in the right hand side panel was generated using 25 different samples of random additive noise selected from a normal distribution with 0 mean 0.05 standard deviation. The value used for standard deviation reflects a 5\% noise level. The average and standard deviation values shown for identifiability of parameters with data in the presence of noise were calculated on the basis of the number of data combinations that can identify the respective parameter for each of the 25 noisy data samples. As seen in these figures, irrespective of the presence or absence of noise in the data, the number of combinations that can identify each parameter, and concordantly, the degree of identifiability of each parameter varies widely. Next, we look at some of the trends in the degree of identifiability of each parameter in each flux.

Since both fluxes $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law, we compare the number of data combinations that are capable of identifying the different parameters used to model these fluxes. The number of data combinations identifying the maximum reaction rates ($V_i^{max}$) for both $v_1$ (Figure \ref{fig:figure1}a and d) and $v_2$ (Figure \ref{fig:figure1}b and e) is always greater than the number of data combinations that identify the enzyme binding ($K_i$) constants. This trend is observed for data combinations with and without noise. Even for parameters in $v_3$ the percentage of data combinations identifying $V_3^{max} (1)$ and $V_3^{max}(2)$ are greater than the percentage of combinations that can identify $K_3^{fdp}(1)$ and $K_3^{fdp}(2)$ (Figure \ref{fig:figure1}c and f). Hence, all maximum reaction rate parameters in the network have a higher degree of identifiability that their corresponding enzyme binding constants.

The difference between the degrees of identifiability of each parameter within a single flux varies widely depending on the reaction flux. For instance, the difference between the percentage of combinations that can identify the parameters used to model $v_1$ is less than 2\% (Figure \ref{fig:figure1}a and d). In contrast, a high number of combinations (89.5\%) can identify $V_2^{max}$ in $v_2$, while less than 2\% of the combinations are able to identify $K_2^{pep}$ (Figure \ref{fig:figure1}b and e). This variability in the degree of identifiability for various parameters modeling the same flux is also seen for $v_3$ (Figure \ref{fig:figure1}c and f): $V_3^{max} (1)$ and $V_3^{max} (2)$ have an average degree of 75\% in comparison to the 15\% identifiability for both $K_3^{fdp} (1)$ and $K_3^{fdp} (2)$. Again, this observation is common for both data with and without added noise.

Also notable is the difference in the degree of identifiability between the maximum reaction rates of the uptake flux $v_1$ and the intracellular fluxes $v_2$ and $v_3$. We discuss this difference and its implications for experimental design in the following section.

\subsection{Experiments required for identifying parameters depends on the position of the flux in the metabolic network:}
Only about 20\% of the 4896 different data combinations can identify either the maximum reaction rate ($V_1^{max}$) or the enzyme turnover rate ($k_1^{cat}$) in $v_1$ (Figure \ref{fig:figure1}a and d), in contrast to the high number of data combinations capable of identifying $V_2^{max}$ (89.5\%) (Figure \ref{fig:figure1}b and e), $V_3^{max} (1)$ or $V_3^{max} (2)$ ($>$75\%) (Figure \ref{fig:figure1}c and f).

This informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending the position of the flux in the metabolic network with respect to the inputs and the outputs of the network. The dependency can be further elucidated using the closed form solutions obtained for parameters of $v_1$ in Section \ref{sec:example}. The identifiability expressions for both parameters in $v_1$ ($V_1^{max}$ and $K_1^{ac}$ in Equation \ref{eq:v1_par}) are dependent on changes in the both the uptake flux $v_1$ as well as the acetate concentration $ac$ in both the experiments used for identification. If either of these values cannot be distinguished between the two experiments whose data is used for identification, then the corresponding combination of experiments can be guaranteed to result in practical non-identifiability of the parameters in $v_1$. Furthermore, the distinct values should satisfy the requirements for practical identifiability mentioned earlier for each parameter (Figure \ref{fig:ident-flowchart}b).

As the uptake flux for the metabolic network that is modeled to be dependent on the substrate concentration, the only way meaningful changes in $v_1$ can be observed are through experiments involving changes either in the substrate concentration or changes in the enzyme concentration for $v_1$. As the change in $v_1$ is directly proportional to changes in acetate concentrations, one can except the conditions for practical identifiability of parameters of $v_1$ (Equation \ref{eq:v1_par}) to be satisfied with relative certainty when experimental data gathered under different acetate concentrations is used for identification. This is further illustrated in Figure \ref{fig:figure2} where the different experiments in each combination identifying both $V_1^{max}$ and $k_1^{cat}$ are shown.

However, in certain cases unlike $v_1$, satisfaction of this conditions may not be as easy as one might except due to the nonlinearity in the relationship between different parts of the network.

We see that experiments involving changes to the substrate concentration figure prominently in the data combinations (as experiment 1) used to identify $V_1^{max}$ or $k_1^{cat}$, and the corresponding enzyme binding constant $K_1^{ac}$ (see Supplementary Figure S1). Experiments involving perturbations to other parameters appear in higher percentages as experiment 2 in the combination identifying these parameters. This can be attributed to the level of change in both $v_1$ as well a

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v1max_k1cat_all_experiments_jan01}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_1^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination. Different experiments that contribute towards identifiability of $k_1^{cat}$ c) experiment 1 ($k=1$) and d) experiment 2 ($k=2$) in combination.}\label{fig:figure2}
\end{figure}

In contrast, any experiment that can distinguish between different values of $pep$ and $v_2$ can be used to identify $V_2^{max}$.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v2max_k2pep_no_noise_experiments_jan03_ab}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_2^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination. Different experiments that contribute towards identifiability of $K_2^{pep}$ c) experiment 1 ($k=1$) and d) experiment 2 ($k=2$) in combination.}\label{fig:figure3}
\end{figure}

When the experimental data has no measurement noise associated with it, then $V_2^{max}$ is the only parameter that can be identified by the most number of experimental combinations (89.5\%). In contrast to the high identifiability of this parameter, $K_2^{pep}$, the other parameter for $v_2$ is identified by less than 2\% of the available data combinations.

This is followed by $V_3^{max}$ that is capable of being identified by 76.3\% of the available data combinations.

In the small model that we use to elucidate our methodology, we find that only

\subsection{Utility of data for identifiability purposes:}

\subsection{Experiments that can establish identifiability for each flux in Kotte model:}	
As indicated previously in Section, we performed different experiments by perturbing the parameters for fluxes $v_1$, $v_2$ and $v_3$ in the model. We also perturbed the input substrate concentration to collect data on the concentrations and fluxes within the network. We describe the results of using this data to establish the identifiability of various parameters in the kinetic model of the small metabolic network.

>>>>

\begin{document}
\section{Introduction:}
Kinetic models of metabolism can be used to study the dynamic characteristics of metabolic networks. In these models, ordinary differential equations (ode) are used to express the rate of change of metabolite concentrations ($x$) as a function of the reaction fluxes ($v$) in the metabolic network (Equation \ref{eq:kinstoich}). The matrix $\mathbf{S}$ in Equation (\ref{eq:kinstoich}a) defines the stoichiometric relationship between the fluxes and the concentrations of the metabolic network.
\begin{subequations}\label{eq:kinstoich}
	\begin{align}
	\dot{x} = \mathbf{S}v\\
	v = f(x, \theta, u)
	\end{align}
\end{subequations}
The expression for the nonlinear function ($f$) used to describe each reaction flux $v_i$ in a kinetic model (Equation \ref{eq:kinstoich}b) is dependent on the enzyme kinetic mechanism that is used to model the reaction \parencite{Heijnen2005, Link2014, Machado2011, Srinivasan2015}. Accordingly, $f$ is a nonlinear function of the metabolite concentrations, enzyme kinetic parameters ($\theta$) and other input concentrations ($u$).

@@

In this paper, we propose a scalable methodology to establish practical identifiability for parameters in kinetic models of metabolism using abundantly available steady state concentration and flux data. We present a computer algebra-based method that can facilitate experimental design through practical identifiability of parameters separately for each individual reaction within a metabolic network based on available steady state experimental data. We illustrate the utility of this method by applying it to a small network of gluconeogenesis in \textit{E. coli} and demonstrating our ability to propose experiments that will facilitate parameter estimation for a kinetic model of this network. %We also demonstrate the scalability of the proposed methodology to facilitate experimental design by applying it to a relatively larger metabolic network of the human red blood cell hepatocyte.

@@

The minimum number of experiments from which data is required for identifying all the parameters of a given flux is determined by dimension $\mathbb{R}^p$ of the parameter space of a chosen flux $v_i$. For instance, as demonstrated above, data from two distinct experiments is required for identifying the two parameters of $v_1$. In this case $p=2$. This also applies for identifying parameters in $v_2$. Hence, multiple data sets generated using a combination of any two experiments are used to test the identifiability of parameters for $v_1$ and $v_2$. The total number of such possible combinations is 306 (18 x 17) from the 18 different experiments mentioned earlier. A similar calculation for identifying the three parameters of $v_3$, where $p=3$, leads to 4896 distinct data combinations (18 x 17 x 16) of three experiments. 
% 18 different experiments mentioned earlier and elaborated in the Appendix.

\subsection{Degee of Identifiability, a quantitative measure of practical identifiability of parameters:}\label{sec:degree_of_identifiability}
We express the practical identifiability of kinetic parameters using a simple quantitative term called the degree of identifiability. We describe the degree of identifiability of any single parameter as the percentage of all data combinations (used to test practical identifiability) that can identify that parameter. As an example, if 90\% of all the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%. Furthermore, we can create a hierarchy of practically identifiable parameters using their degree of identifiabilities. In the above instance of the two parameters $\theta_i$ and $\theta_j$ that have degrees of identifiabilities of 90\% and 50\% respectively, $\theta_i$ is classified to be more identifiable than $\theta_j$ due to its relatively higher degree of identifiability. Determining this hierarchy of identifiable parameters can help in distinguishing parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments, and design minimal sets of experiments that can practically identify all kinetic parameters used to model a metabolic network.

\subsection{Experimental design through practical parameter identifiability for kinetic models of metabolism:}\label{sec:experimental_design}
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure3/experimental_design}}
	\caption{Flow diagram showing a method for experimental design that uses our methodology for practical identifiability of parameters to determine the number and type of experiments required to identify all fluxes within a given metabolic network.}\label{fig:ident-design}
\end{figure}
Following the methodology described in Section \ref{sec:ident}, and demonstrated in Section \ref{sec:example} for a single flux using data from a combination of two different experiments, all distinct combinations found from Section \ref{sec:experiments} can be tested for their ability to practically identify parameters for any of the three fluxes in the small metabolic network. This step would help distinguish identifiable experiment combinations from combinations that do not practically identify any parameter in the model (Figure).
Subsequently, it is possible to obtain a collection of experiments that make up all identifiable data combinations that can be performed to obtain the most minimal and informative set of experiments to identify as many parameters as possible (Figure \ref{fig:ident-design}). Consequently, the set of experiments can be used to estimate the identifiable parameters in the model. We explain this formally below.

The identifiability of each parameter based on each experiment indexed as $j = \{1, ..., n\}$ is established based on the methodology described previously in Section \ref{sec:ident} and demonstrated in Section \ref{sec:example}. Subsequently, for any flux $v_i$, if for any $p$ combinations of indices $j$, the experimental concentrations ($\mathbf{x}_j$) and fluxes ($\mathbf{v}_j$) do not satisfy the condition for identifiability for any parameter in $\theta\in\mathbb{R}^p$, i.e., $D_k(\mathbf{v}_i, \mathbf{x}, \mathbf{u}) = 0$ for any $k$, then at least one of the $p$ experiments needs to be changed to make a parameter $\theta_k$ identifiable. Consequently, the corresponding experiment cannot be used for parameter estimation and needs to be discarded from the set of all necessary experiments. Furthermore, another experiment from $j = \{1, ..., n\}$ needs to be selected such that parameter $\theta_k$ is identifiable. This process has to be repeated until all parameters in $\theta\in\mathbb{R}^p$ are identifiable for flux $v_i$. In doing so, we can arrive at a set of $p$ experiments that will always result in practically identifiable parameters for flux $v_i$. This analysis can be performed for each flux in a metabolic network independent of all the other fluxes. Hence, our method is theoretically scalable even to genome-scale models.

\section{Results:}
First in Section \ref{sec:closedform}, we show that the ability to establish practical identifiability of kinetic parameters in metabolic network models using our methodology, described in Section \ref{sec:ident}, relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We use the gluconeogenic model described in Section \ref{sec:example} as an example. Then, in Section \ref{sec:trends} we show how the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetics rate law used to describe the flux. In Section \ref{sec:experiment_type} that follows, we discuss the the ability to determine experiments for parameter identifiability for a given flux based on the informativeness of a given experiment. In this section, we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identifiability of the parameters for a given flux. In Section \ref{sec:noise} we discuss the identifiability of parameters when data with additive noise is used with our methodology to test practical identifiability. Finally, in Section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models metabolic networks.

\subsection{Nonlinearity of enzyme kinetic rate law expression affects identifiability analysis:}\label{sec:closedform}
Establishing parameter identifiability for fluxes in kinetic models of metabolism using the methodology we describe in the Methods section is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. We demonstrate one example of how the methodology works in Section \ref{sec:example} for parameters of flux $v_1$. The expression in Equation (\ref{eq:v1_par}) is obtained by using a computer algebra system. To recall, in this specific case, a computer algebra system is used to solve for the parameters of a flux described using the Michaelis-Menten kinetic rate law when experimental data from two different experiments is available.

However, we find that the nonlinearity of the MWC kinetic rate law used to model the allosteric regulation of $v_3$ makes it computationally intractable for determining the closed form expressions of the three parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ using a computer algebra system (Mathematica and SymPy in Python). In order to overcome this computational obstacle, we model the reaction rate for $v_3$ using the convenience kinetic rate law formulation \parencite{Liebermeister2006}. The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}).

@@

The impact of nonlinearity of a kinetic rate law expression affecting the practical identifiability of parameters in the expression may not be an unique problem isolated to the system that we are investigating. Complicated expressions for describing fluxes have been extensively used to model observed experimental data for different fluxes in a variety of organisms \parencite{Chassagnole2002a, Peskov2012, VanHeerden2014}. However, the identifiability of the parameters used in these models has never been truly examined. If these models are tractable, we believe our methodology can help in elucidating the identifiability of parameters in these models. If not, metabolic network fluxes can be expressed using alternative kinetic rate law models whose parameters can be tested for identifiability, and subsequently experiments can be designed for their estimation and model validation.

Next, we look some of the results we obtained for the practical identifiability of different parameters for all three fluxes $v_1$, $v_2$ and $v_3$ in the small gluconeogenic network model described in Section \ref{sec:small-model}, with the flux for $v_3$ described by Equation (\ref{eq:flux3_convkin}).

\subsection{Maximum reaction rates are more identifiable than enzyme binding constants:}\label{sec:trends}
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/v1_v2_v3_identifiability_no_noise}}
	\caption{The number of data combination from 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data is shown. The percentage of total experimental data combinations that can identify each parameter is specified in/on the right hand side of the bar showing the total number of combinations identifying a given parameter. The total number of data combinations used to identify parameters in fluxes $v_1$ and $v_2$ is lower that the total number of combinations used to test identifiability of parameters for $v_3$ since both $v_1$ and $v_2$ have only two parameters while $v_3$ is modeled with three parameters. Hence, the total number of experiments used to identify parameters in $v_1$ and $v_2$ is two, while parameter identification in $v_3$ requires data from at least three different experiments.}\label{fig:figure1}
\end{figure}
In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. Based on the definition given in Section \ref{sec:degree_of_identifiability}, the percentages refer to the degree of identifiability of each parameter. The three panels in Figure \ref{fig:figure1} represent the degree of identifiability for parameters modeling the three different fluxes of the small network individually. Looking at these panels for $v_1$ (Figure \ref{fig:figure1}a), $v_2$ (Figure \ref{fig:figure1}b) and $v_3$ (Figure \ref{fig:figure1}c), the degree of identifiability of the maximum reaction rates ($V_i^{max}$) in each of the three fluxes is higher than the degree of identifiability of the enzyme binding ($K_i$) constants and the allosteric activation constant ($K_3^{pep}$). We observe this trend irrespective of the fact that only  $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law (Michaelis-Menten), while the convenience kinetic rate law is used to model $v_3$. These observations lead us to conclude that the maximum reaction rate parameters are always more identifiable (as indicated by the higher degree of identifiability) than their enzyme binding constant counterparts irrespective of the enzyme kinetic rate law used to model the corresponding flux.

The most conspicuous of all observations from Figure \ref{fig:figure1} is the difference in the identifiability of parameters based on the flux in which these parameters are present. The variation in the degree of identifiability is not only dependent on the type of parameter being identified, but also depends on the  flux whose parameters are being tested for identifiability. This can be tied to selecting experimental data that have the ability to satisfy conditions for practical identifiability for a given parameter. We discuss the dependence of the variation in the degree of identifiability for parameters of different fluxes from the point of view of the informativeness of the experiment from which data is collected for identifiability in the following section.

\subsection{Experiments required for identifying parameters depends on the position of the flux in the metabolic network:}\label{sec:experiment_type}
In system identification, any input signal should be rich or informative enough to guarantee full excitement of the dynamics of the system. Only information obtained from such changes in the input and the corresponding signals can be used to completely identify the system over its entire dynamic range. Input signals that guarantee to fully excite the dynamics of the system are termed as persistently excitable signals. In linear systems, persistence of excitation of any input signal is guaranteed. However, for nonlinear systems, like the metabolic network we deal with in this paper, persistence of excitation of input signals is not guaranteed and should be assessed on a case by case basis. The lack of persistence of excitation requires the design of experiments to satisfy the data needs for complete system identification.

In this section, we demonstrate a specific case of lack of persistence of excitation and the subsequent experimental design required to overcome this hurdle for parameter identifiability for a small nonlinear metabolic network model. The informativeness of experiments for identifying parameters in a given flux depends on its ability to satisfy the conditions determined for practical identifiability of that parameter. For example, we demonstrated in Section \ref{sec:example} that for a combination of any two experiments to be capable of identifying $V_1^{max}$ and $K_1^{ac}$ in $v_1$, the experiments must have distinct acetate concentrations as well as a different uptake flux $v_1$ between them. Thus, in this instance, the informativeness of the experiments (changes in the input acetate concentration) for identifying parameters of $v_1$ is determined by their ability to change the measured value of flux $v_1$. Similarly, the identification of parameters for $v_2$ requires the experiments to distinguish between values of both $v_2$ as well as \textit{pep}, given both these variables are outputs of the model describing $v_2$.

In Figure \ref{fig:figure1}a we show that only about 20\% of all the experimental data sets satisfy these requirements for $v_1$ while more than 90\% of the available experimental satisfies these requires for $v_2$ (Figure \ref{fig:figure1}b). Thus, experimental design becomes crucial to determine the minimum number, as well as the nature of experiments that can help identify parameters for both $v_1$ and $v_2$. Recall that we use experimental data from five different experiments to test the practical identifiability of parameters in the model (Section \ref{sec:experiments}).
In order to determine the type of experiments that help satisfy identifiability conditions for both parameters of $v_1$, we look at the distribution of the five different experiments in all data combinations that can identify these 2 parameters (Figure \ref{fig:figure2}).

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v1max_k1ac_experiments}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_1^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination, and $K_1^{ac}$ c) experiment 1 ($k=1$), and d) experiment 2 ($k=2$) in combination. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:figure2}
\end{figure}

The contribution of experiments that involve changes in the acetate concentrations, which consequently bring about changes in the value of $v_1$ contribute to a significant part (> 50\%) of the identifiable experimental data combinations in contrast to the other types of experiments. This matches with requirements for identifiability on the basis of the informativeness of experiments that was laid out earlier i.e., experiments must be able distinguish between changes in $v_1$ on the basis of changes in acetate. This informs why only about 20\% of the 306 different experiment combinations can identify these parameters (Figure \ref{fig:figure1}a). In can be gathered that only about 20\% of all the data combinations used to test identifiability are informative enough to distinguish between both acetate as well as $v_1$, and most of these experiments require changes in the acetate concentration to be able to identify $V_1^{max}$ and $K_1^{ac}$ through changes in the input and the output quantities acetate and $v_1$, respectively.

In contrast to $v_1$, we see that the different perturbation experiments have a higher contribution towards experimental data combinations that can identify parameters for $v_2$ (Figure). Also, it is important to note that since most experiment types can bring about changes in the dynamics of $v_2$ (changes to \textit{pep} and value of $v_2$), the identifiability of parameters for $v_2$ is relatively higher than the identifiability of $v_1$ parameters (Figure \ref{fig:figure1}b). We can further extend these arguments to also justify the observed contribution of experiments towards identifying parameters for $v_3$ (figure not shown).

Alternatively, both the degree of identifiability and the capability of different experiments to identify parameters of different fluxes based on their informativeness can be explained by the position of the flux in the metabolic network. We believe that the position of any given in the metabolic network determines the specific experiment that is persistently excitable enough to identify the parameters of that flux. This dependency of experiment informativeness on the position of the flux can be further elucidated using $v_1$ and $v_2$ as examples. We know from Equation (\ref{eq:v1_par}) that identifiability of $v_1$ requires changes in both acetate and $v_1$. We also know, based on our knowledge of the Michaelis-Menten kinetic rate law that changes in the substrate concentration of a reaction can bring about a nonlinear change in the value of the corresponding reaction rate (flux). In this specific metabolic network, since the substrate is an input variable to the model and $v_1$ is the corresponding uptake flux, the substrate can be easily perturbed to create persistently excitable experiments to identify parameters in $v_1$. We can generalize this observation for the identification of all uptake fluxes in all metabolic networks, i.e., at a minimum, a change in the input substrate concentration may be necessary for an informative experiment to identify the uptake flux parameters.

On the other hand, the Michaelis-Menten model for $v_2$ also requires changes in \textit{pep} and $v_2$ for persistently excitable experiments to identify $v_2$. However, since both of these are system outputs, satisfaction of this condition cannot be guaranteed without an analysis of the dynamics of the metabolic network and how changes in the input (acetate) bring about changes in the two requisite output quantities. Previously, a dynamical analysis of the current network has shown that the concentration of \textit{pep}, and hence the flux $v_2$ are functions of not only the input acetate concentration, but also the levels of expression of the different enzymes within the network (Srinivasan et al., 2017). Hence, it is theoretically possible for any of the five different experiment types to be persistently excitable to identify $v_2$. This is confirmed by the high degree of identifiability for both parameters of $v_2$, where in more than 90\% of all data combinations can identify the parameters.
Thus, this analysis informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending the position of the flux in the metabolic network with respect to the inputs and the outputs of the network.

Nonetheless, all these explanations could only be verified due to the fact that the conditions for identifiability for both $v_1$ and $v_2$ are simple and readable due to the use of the Michaelis-Menten rate law to model these fluxes. Instead, in addition to presence of an allosteric interaction, when the convenience kinetic rate law is used to model this activation in $v_3$, we find that the complex expressions obtained for the three parameters preclude such an in-depth insight into the nature of persistently excitable experiments to identify  $v_3$. Thus, the only avenue through which experiments can be designed for the identification of parameters of $v_3$ would be through our proposed methodology wherein manual curation of experimental combinations required for identifiability are not necessary.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v2max_k2pep_experiments}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_2^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination. Different experiments that contribute towards identifiability of $K_2^{pep}$ c) experiment 1 ($k=1$) and d) experiment 2 ($k=2$) in combination.}\label{fig:figure3}
\end{figure}
